---
title: "Cianini_bulk"
author: "Eleonora Cianini"
date: "9/30/2021"
output: html_document
---

# [I] - Libraries
Load all the libraries required for the analysis
```{r, message=FALSE}
library(recount)
library(limma)
library(edgeR)
library(ggplot2)
library(AnnotationDbi)
library(clusterProfiler)
library(org.Hs.eg.db)
library(xlsx)
```


# [II] -Retrieving the data and Pre-processing the count tables
All the data comes from the GTEx project database, and here we got the Liver, Heart and Colon's tissues, with 3 samples per tissue: 

## Liver
```{r, message=FALSE}
#####LIVER#####
load("rse_gene_liver_9_scaled.Rdata")
liver_rse <- rse
```

##### 1. Select the three columns in which we have to work on
```{r, message=FALSE}
liver_count <- assays(liver_rse)$counts[, 3:5]
dim(liver_count)
tot_reads_liver <- colSums(liver_count)
```

##### 2. Remove all genes with length < 200
```{r, message=FALSE}
small_genes_9 <- rowRanges(liver_rse)[rowData(liver_rse)$bp_length < 200, ]
small_liver <-which(rownames(liver_count) %in% small_genes_9$gene_id)
clean_small_9 <- liver_count[-c(small_liver),] 
cleaned_small_9 <- colSums(clean_small_9) 

## produce the overall read count associated with these genes in each tissue/replicate
short_RNAs_liver <- (tot_reads_liver - cleaned_small_9) #73154 47129 73677
perc_short_RNAs_liver <- (short_RNAs_liver/tot_reads_liver)*100
```

##### 3. Remove all mitochondrial genes
```{r, message=FALSE}
mit_genes_9 <- rowRanges(liver_rse)[seqnames(rowRanges(liver_rse)) == "chrM" &
                                      rowRanges(liver_rse)$bp_length >= 200, ]

#15 mitocondrial genes (22 are mitocondrial genes < 200bp counted in the small)
mit_liver <-which(rownames(liver_count) %in% mit_genes_9$gene_id)
clean_mit_9 <- liver_count[-c(mit_liver),] 
cleaned_mit_9 <- colSums(clean_mit_9) 

## produce the overall read count associated with these genes in each tissue replicate
mit_RNAs_liver <- (tot_reads_liver - cleaned_mit_9) 
perc_mit_RNAs_liver <- (mit_RNAs_liver/tot_reads_liver)*100

clean_liver <- liver_count[-c(small_liver, mit_liver),]
dim(clean_liver)
```

## Heart
```{r, message=FALSE}
#####HEART#####
load("rse_gene_heart_6_scaled.Rdata")
heart_rse <- rse
```


##### 1. Select the three columns in which we have to work on
```{r, message=FALSE}
heart_count <- assays(heart_rse)$counts[, 2:4] 
tot_reads_heart <- colSums(heart_count) 
```

##### 2. Remove all genes with length < 200
```{r, message=FALSE}
small_genes_6 <- rowRanges(heart_rse)[rowRanges(heart_rse)$bp_length < 200, ]
small_heart <-which(rownames(heart_count) %in% small_genes_6$gene_id)
clean_small_6 <- heart_count[-c(small_heart),] 
cleaned_small_6 <- colSums(clean_small_6) 

## produce the overall read count associated with these genes in each tissue/replicate
short_RNAs_heart <- (tot_reads_heart - cleaned_small_6) 
perc_short_RNAs_heart <- (short_RNAs_heart/tot_reads_heart)*100
```


#####3. Remove all mitochondrial genes
```{r, message=FALSE}
mit_genes_6 <- rowRanges(heart_rse)[seqnames(rowRanges(heart_rse)) == "chrM" &
                                      rowRanges(heart_rse)$bp_length >= 200, ]
mit_heart <-which(rownames(heart_count) %in% mit_genes_6$gene_id)
clean_mit_6 <- heart_count[-c(mit_heart),]
cleaned_mit_6 <- colSums(clean_mit_6) 
```
Produce the overall read count associated with these genes in each tissue replicate
```{r, message=FALSE}
mit_RNAs_heart <- (tot_reads_heart - cleaned_mit_6)   
perc_mit_RNAs_heart <- (mit_RNAs_heart/tot_reads_heart)*100

clean_heart <- heart_count[-c(small_heart, mit_heart),]
dim(clean_heart)
```

## Colon
```{r, message=FALSE}
#####COLON#####
load("rse_gene_colon_7_scaled.Rdata")
colon_rse <- rse
```


##### 1. Select the three columns in which we have to work on
```{r, message=FALSE}
colon_count <- assays(colon_rse)$counts[, c(9, 10, 1)] # 58037 genes & 3 samples
tot_reads_colon <- colSums(colon_count)# 38364393   38984105   37414767
```
##### 2. Remove all genes with length < 200
```{r, message=FALSE}

small_genes_7 <- rowRanges(colon_rse)[rowRanges(colon_rse)$bp_length < 200, ]
small_colon <-which(rownames(colon_count) %in% small_genes_7$gene_id)
clean_small_7 <- colon_count[-c(small_colon),] 
cleaned_small_7 <- colSums(clean_small_7) 
#produce the overall read count associated with these genes in each tissue/replicate
short_RNAs_colon <- (tot_reads_colon - cleaned_small_7) 
perc_short_RNAs_colon <- (short_RNAs_colon/tot_reads_colon)*100
```

#####3. Remove all mitochondrial genes
```{r, message=FALSE}
mit_genes_7 <- rowRanges(colon_rse)[seqnames(rowRanges(colon_rse)) == "chrM" &
                                      rowRanges(colon_rse)$bp_length >= 200, ]
#15 mitocondrial genes (22 are mitocondrial genes < 200bp counted in the small)
mit_colon <-which(rownames(colon_count) %in% mit_genes_7$gene_id)
clean_mit_7 <- colon_count[-c(mit_colon),]
## produce the overall read count associated with these genes in each tissue replicate
cleaned_mit_7 <- colSums(clean_mit_7)

## produce the overall read count associated with these genes in each tissue replicate
mit_RNAs_colon <- (tot_reads_colon - cleaned_mit_7)
perc_mit_RNAs_colon <- (mit_RNAs_colon/tot_reads_colon)*100

clean_colon <- colon_count[-c(small_colon, mit_colon),]
dim(clean_colon)
```


# [III] - Merging the data
Obtained all the information required from the count tables downloaded, we merged all the elements together a DGEobject, which contains all the information for the downstream analysis.

```{r}
#Merge the data
data <- (cbind(clean_liver, clean_heart, clean_colon))
colSums(data)

#####DE analysis#####
dim(data) #50681 genes 9 replicates

#create DGElist
y <- DGEList(counts=data)
head(y)
#label the samples
group <- as.factor(c("liver", "liver", "liver", 
                     "heart", "heart", "heart", 
                     "colon", "colon", "colon"))
y$samples$group <- group
head(y)
```

# [IV] - Filtering step
An important step for a good analysis is to obtain the counts filtered from the low-expressed genes.

```{r}
#REMOVE the low-expressed genes 
keep.exprs <- filterByExpr(y, group=group)
# filterByExpr remove genes that are not expressed
## above a log-CPM threshold in a minimum number of samples

summary(keep.exprs)

y <- y[keep.exprs,, keep.lib.sizes=FALSE]

# Almost 27k  genes have been filtered out with default parameters.
dim(y)
# The remained genes are 23786

# Store the non-normalized counts 
logcpm_before <- cpm(y, log=TRUE)
```

# [V] - Normalization step
```{r}
# Normalize the counts (TMM normalization, trimmed mean)
y <- calcNormFactors(y, method = "TMM")
#y$samples

logcpm <- cpm(y, log=TRUE)
```

```{r}
# boxplot of the non-normalized counts
x11()
boxplot(logcpm_before, notch=TRUE, outline = FALSE,
        col=(c(rep("lightcyan3",3),rep("lightgoldenrod3", 3), 
               rep("lightpink3",3))),
        ylim = c(-5,20),
        main="Log2CPM not normalized", xlab="Sample")
legend(x = "topright",
       legend = c("Liver","Heart","Colon"),
       fill =c("lightcyan3","lightgoldenrod3",
                 "lightpink3")
)

graphics.off()
```

```{r}
# boxplot of the normalized counts
x11()
boxplot(logcpm, notch=TRUE, outline = FALSE,
        col=(c(rep("lightcyan3",3),rep("lightgoldenrod3", 3), 
               rep("lightpink3",3))),
        ylim = c(-5,20),
        main="Log2CPM post TMM normalization", xlab="Sample")
legend(x = "topright",
       legend = c("Liver","Heart","Colon"),
       fill =c("lightcyan3","lightgoldenrod3",
               "lightpink3"),
       ncol = 3
)

graphics.off()
```

By eyes we can see a change in the boxplot of the normalized counts if compared to the non_normalized, in fact after the normalization more of the samples are closer to the median horizontal line. Normalized the data we can start with the desgin model.
```{r}
#design the linear model
design <- model.matrix(~0+group, data=y$samples)
# no intercept is used
colnames(design) <- levels(y$samples$group)
design
```

# [VI] - Exploring data and analyzing them
```{r}
# plot the samples labeled using MDS by tissue type
x11()
plotMDS(cpm(y, log=TRUE), 
        labels = colnames(logcpm),
        col = c(rep("lightcyan3",3), rep("lightgoldenrod3", 3),rep("lightpink3", 3)),
        cex = 2
        )
legend("topright", legend = c("Liver", "Heart", "Colon"),
       fill = c("lightcyan3", "lightgoldenrod3", "lightpink3" ))
title(main = "MDS plot of the log-normalized samples")

graphics.off()

# plot the samples labeled by gender
gender <- as.factor(c("male", "male", "male", 
                      "female", "male", "male", 
                      "female", "male", "male"))
y$samples$gender <- gender
y

x11()
plotMDS(cpm(y, log=TRUE), 
        labels = colnames(logcpm),
        col=ifelse(y$samples$gender =="female", "lightpink2", "mediumblue"),
        cex = 2)
legend("topright", legend = c("Female", "Male"),
       fill = c("lightpink2", "mediumblue"))
title(main = "MDS plot of the log-normalized samples by gender")

graphics.off()
```

Observing the data, we investigate if the age 60-69 could be a biological factor or not. 
```{r}
# plot the samples labeled by age bracket
age <- as.factor(c("30-39", "60-69", "60-69", 
                      "20-29", "60-69", "60-69", 
                      "40-49", "40-49", "50-59"))
y$samples$age <- age
#y

x11()
plotMDS(cpm(y, log=TRUE), 
        labels = colnames(logcpm),
        col=ifelse(y$samples$age =="60-69", "palegreen3", "sienna3")
)
legend(2, 4, legend = c("Age 60-69", "Other age ranges"),
       fill = c("palegreen3", "sienna3"))

title(main = "MDS plot of the log-normalized samples by age bracket")
graphics.off()

# Estimate BCV
# We estimate the negative binomial (NB) dispersions with estimateDisp()
y <- estimateDisp(y, design)
summary(y$trended.dispersion)
#y$common.dispersion
```

```{r}
x11()
plotBCV(y)
title(main = "BCV and average logCPM plot")

graphics.off()
```

```{r}
# fit the data to the designed generalized linear model
fit <- glmQLFit(y, design)

# COLON VERSUS HEART
# 1vs2: colon versus heart
qlf.1vs2 <- glmQLFTest(fit, contrast = c(1, -1, 0))

# select the significant ones, with corrected p-value (FDR) < 0.05: colon
# versus colon
FDR.1vs2 <- p.adjust(qlf.1vs2$table$PValue, method = "BH")

#append the adjusted pvalue in the table
qlf.1vs2$table$FDR <- FDR.1vs2
#qlf.1vs2

#plot 
p <- ggplot(data=qlf.1vs2$table, 
aes(x=logFC, y=-log10(FDR))) + geom_point() +
  theme_minimal() +
  scale_x_continuous(breaks=seq(-15, 15, 0.5)) +
  scale_y_continuous(breaks=seq(0, 3.5, 0.5))+
  ggtitle("COLON vs HEART")
p

p2 <- p + geom_hline(yintercept=-log10(0.05), col="red")
p2

graphics.off()

# add a column of NAs
qlf.1vs2$table$DEgenes <- "NO"
# if log2Foldchange > 0 and pvalue < 0.05, set as "UP" 
qlf.1vs2$table$DEgenes[qlf.1vs2$table$logFC > 1 & qlf.1vs2$table$FDR < 0.05] <- "UP"
# if log2Foldchange < 0 and pvalue < 0.05, set as "DOWN"
qlf.1vs2$table$DEgenes[qlf.1vs2$table$logFC < -1 & qlf.1vs2$table$FDR < 0.05] <- "DOWN"

# Re-plot but this time color the points with "DEgenes"
p <- ggplot(data=qlf.1vs2$table, 
            aes(x=logFC, y=-log10(FDR), 
                col=DEgenes)) +
  scale_x_continuous(breaks=seq(-15, 15, 0.5)) +
  scale_y_continuous(breaks=seq(0, 3.5, 0.5)) + 
  geom_point() + 
  theme_minimal()+
  ggtitle("COLON vs HEART")

# Add lines as before...
p2 <- p + geom_vline(xintercept=c(-1.25, 1.25), col="red") +
  geom_hline(yintercept=-log10(0.05), col="red") 
  
mycolors <- c("turquoise3", "violetred3", "black")
names(mycolors) <- c("DOWN", "UP", "NO")
p3 <- p2 + scale_colour_manual(values = mycolors)
p3
graphics.off()

summary(decideTests(qlf.1vs2, p.value = 0.05))

deg.1vs2 <- topTags(qlf.1vs2, n = 20000, adjust.method = "BH", sort.by = "PValue",
                    p.value = 0.05)$table


up.genes.1vs2 <- row.names(deg.1vs2[deg.1vs2$logFC > 1, ]) #233
down.genes.1vs2 <- row.names(deg.1vs2[deg.1vs2$logFC < -1, ]) #573 
#deg.1vs2
```

Usually, the transformation is log at base 2, so the interpretation is straightforward: a log(FC) of 1 means twice as expressed.

```{r}
#COLON vs LIVER
qlf.1vs3 <- glmQLFTest(fit, contrast = c(1, 0, -1))

# select the significant ones, with corrected p-value (FDR) < 0.05: heart
# versus colon
FDR.1vs3 <- p.adjust(qlf.1vs3$table$PValue, method = "BH")

#append the adjusted pvalue in the table
qlf.1vs3$table$FDR <- FDR.1vs3
#qlf.1vs3

#plot 
p <- ggplot(data=qlf.1vs3$table, 
            aes(x=logFC, y=-log10(FDR))) + geom_point() +
  theme_minimal() +
  scale_x_continuous(breaks=seq(-15, 15, 0.5)) +
  scale_y_continuous(breaks=seq(0, 3.5, 0.5))+
  ggtitle("COLON vs LIVER")
p

p2 <- p + geom_hline(yintercept=-log10(0.05), col="red")
p2
graphics.off()

# add a column of NAs
qlf.1vs3$table$DEgenes <- "NO"
# if log2Foldchange > 0 and pvalue < 0.05, set as "UP" 
qlf.1vs3$table$DEgenes[qlf.1vs3$table$logFC > 0.75 & qlf.1vs3$table$FDR < 0.05] <- "UP"
# if log2Foldchange < 0 and pvalue < 0.05, set as "DOWN"
qlf.1vs3$table$DEgenes[qlf.1vs3$table$logFC < -0.75 & qlf.1vs3$table$FDR < 0.05] <- "DOWN"

# Re-plot but this time color the points with "DEgenes"
p <- ggplot(data=qlf.1vs3$table, 
            aes(x=logFC, y=-log10(FDR), 
                col=DEgenes)) +
  scale_x_continuous(breaks=seq(-15, 15, 0.5)) +
  scale_y_continuous(breaks=seq(0, 3.5, 0.5)) + 
  geom_point() + 
  theme_minimal()+
  ggtitle("COLON vs LIVER")

# Add lines as before...
p2 <- p + geom_vline(xintercept=c(-0.75, 0.75), col="red") +
  geom_hline(yintercept=-log10(0.05), col="red") 

mycolors <- c("turquoise3", "violetred3", "black")
names(mycolors) <- c("DOWN", "UP", "NO")
p3 <- p2 + scale_colour_manual(values = mycolors)
p3

graphics.off()

summary(decideTests(qlf.1vs3, p.value = 0.05))

deg.1vs3 <- topTags(qlf.1vs3, n = 20000, adjust.method = "BH", sort.by = "PValue",
                    p.value = 0.05)$table
up.genes.1vs3 <- row.names(deg.1vs3[deg.1vs3$logFC >1, ]) #1967 
down.genes.1vs3 <- row.names(deg.1vs3[deg.1vs3$logFC < -1, ])#2210 
#deg.1vs3

#HEART vs LIVER
qlf.2vs3 <- glmQLFTest(fit, contrast = c(0, 1, -1))
FDR.2vs3 <- p.adjust(qlf.2vs3$table$PValue, method = "BH")

#append the adjusted pvalue in the table
qlf.2vs3$table$FDR <- FDR.2vs3
#qlf.2vs3

#plot 
p <- ggplot(data=qlf.2vs3$table, 
            aes(x=logFC, y=-log10(FDR))) + geom_point() +
  theme_minimal() +
  scale_x_continuous(breaks=seq(-15, 15, 0.5)) +
  scale_y_continuous(breaks=seq(0, 3.5, 0.5))+
  ggtitle("HEART vs LIVER")
p

p2 <- p + geom_hline(yintercept=-log10(0.05), col="red")
p2

graphics.off()

# add a column of NAs
qlf.2vs3$table$DEgenes <- "NO"
# if log2Foldchange > 0 and pvalue < 0.05, set as "UP" 
qlf.2vs3$table$DEgenes[qlf.2vs3$table$logFC > 0.75 & qlf.2vs3$table$FDR < 0.05] <- "UP"
# if log2Foldchange < 0 and pvalue < 0.05, set as "DOWN"
qlf.2vs3$table$DEgenes[qlf.2vs3$table$logFC < -0.75 & qlf.2vs3$table$FDR < 0.05] <- "DOWN"

# Re-plot but this time color the points with "DEgenes"
p <- ggplot(data=qlf.2vs3$table, 
            aes(x=logFC, y=-log10(FDR), 
                col=DEgenes)) +
  scale_x_continuous(breaks=seq(-15, 15, 0.5)) +
  scale_y_continuous(breaks=seq(0, 3.5, 0.5)) + 
  geom_point() + 
  theme_minimal()+
  ggtitle("HEART vs LIVER")

# Add lines as before...
p2 <- p + geom_vline(xintercept=c(-0.75, 0.75), col="red") +
  geom_hline(yintercept=-log10(0.05), col="red") 

mycolors <- c("turquoise3", "violetred3", "black")
names(mycolors) <- c("DOWN", "UP", "NO")
p3 <- p2 + scale_colour_manual(values = mycolors)
p3

graphics.off()

summary(decideTests(qlf.2vs3, p.value = 0.05))

deg.2vs3 <- topTags(qlf.2vs3, n = 20000, adjust.method = "BH", sort.by = "PValue",
                    p.value = 0.05)$table
up.genes.2vs3 <- row.names(deg.2vs3[deg.2vs3$logFC > 1, ])#2501 
down.genes.2vs3 <- row.names(deg.2vs3[deg.2vs3$logFC < -1, ])#2654 
#deg.2vs3
```

And now I'm going to find the up and down regulated genes of my dataset

```{r}
# Colon
up.colon <- intersect(up.genes.1vs2, up.genes.1vs3)
down.colon <- intersect(down.genes.1vs2, down.genes.1vs3)

# Heart
up.heart <- intersect(up.genes.2vs3, down.genes.1vs2)
down.heart <- intersect(down.genes.2vs3, up.genes.1vs2)

# Liver
up.liver <- intersect(down.genes.1vs3, down.genes.2vs3)
down.liver <- intersect(up.genes.1vs3, up.genes.2vs3)

```

# [VII] - Prepare the file for the export to Excel
The following code was founded in GitHub taken from the Daniele Colombo's vignette. 
```{r, message=FALSE}
de.genes.dataframe <- function(col1, col2, column.names) {
  de.genes <- data.frame(col1 = rep(NA, max(sapply(list(col1, col2), length))))
  de.genes[1:length(col1), 1] <- col1
  de.genes[1:length(col2), 2] <- col2
  colnames(de.genes) <- column.names
  
  return(de.genes)
}
```

```{r, message=FALSE}
comparison.1vs2 <- de.genes.dataframe(up.genes.1vs2, down.genes.1vs2, c("up.colon.vs.heart",
                                                                        "down.colon.vs.heart"))
comparison.1vs3 <- de.genes.dataframe(up.genes.1vs3, down.genes.1vs3, c("up.colon.vs.liver",
                                                                        "down.colon.vs.liver"))
comparison.2vs3 <- de.genes.dataframe(up.genes.2vs3, down.genes.2vs3, c("up.heart.vs.liver",
                                                                        "down.heart.vs.liver"))

de.genes.colon <- de.genes.dataframe(up.colon, down.colon, c("up.colon", "down.colon"))
de.genes.heart <- de.genes.dataframe(up.heart, down.heart, c("up.heart", "down.heart"))
de.genes.liver <- de.genes.dataframe(up.liver, down.liver, c("up.liver", "down.liver"))
```

Obtaining the xls files of the DE gene list
```{r, message=FALSE}
write.xlsx(comparison.1vs2, "DE_genes.xlsx", sheetName = "Tissue 1 vs 2", col.names = TRUE,
                 row.names = FALSE, append = FALSE, showNA = FALSE)
write.xlsx(comparison.1vs3, "DE_genes.xlsx", sheetName = "Tissue 1 vs 3", col.names = TRUE,
                 row.names = FALSE, append = TRUE, showNA = FALSE)
write.xlsx(comparison.2vs3, "DE_genes.xlsx", sheetName = "Tissue 2 vs 3", col.names = TRUE,
                 row.names = FALSE, append = TRUE, showNA = FALSE)
write.xlsx(de.genes.colon, "DE_genes.xlsx", sheetName = "DE genes tissue 1",
                 col.names = TRUE, row.names = FALSE, append = TRUE, showNA = FALSE)
write.xlsx(de.genes.heart, "DE_genes.xlsx", sheetName = "DE genes tissue 2",
                 col.names = TRUE, row.names = FALSE, append = TRUE, showNA = FALSE)
write.xlsx(de.genes.liver, "DE_genes.xlsx", sheetName = "DE genes tissue 3",
                 col.names = TRUE, row.names = FALSE, append = TRUE, showNA = FALSE)
```

For the functional enrichment analysis we needed to retrieve the list of gene_symbols using Enrichr, so:

```{r, message=FALSE}
# functional enrichment analysis 
write.gene.symbols <- function(ensembl, file.name) {
  symbols <- mapIds(org.Hs.eg.db, keys = gsub("\\..*", "", ensembl), keytype = "ENSEMBL",
                    column = "SYMBOL", multiVals = "first")
  write.table(symbols[!is.na(symbols)], file.name, col.names = F, row.names = F,
              quote = F)
}

write.gene.symbols(up.colon, "up_colon_symbols.txt")
write.gene.symbols(up.heart, "up_heart_symbols.txt")
write.gene.symbols(up.liver, "up_liver_symbols.txt")

```


```{r}
up_1vs2_FDR <- data.frame(deg.1vs2[deg.1vs2$logFC > 0,]$FDR)
up_1vs2_FDR$ENSG <- row.names(deg.1vs2[deg.1vs2$logFC > 0,])
row.names(up_1vs2_FDR) <- up_1vs2_FDR$ENSG

up_1vs3_FDR <- data.frame(deg.1vs3[deg.1vs3$logFC > 0,]$FDR)
up_1vs3_FDR$ENSG <- row.names(deg.1vs3[deg.1vs3$logFC > 0,])
row.names(up_1vs3_FDR) <- up_1vs3_FDR$ENSG

up_both <- merge(up_1vs2_FDR,up_1vs3_FDR)
sort_up_both <- up_both[order(up_both$deg.1vs2.deg.1vs2.logFC...0....FDR,up_both$deg.1vs3.deg.1vs3.logFC...0....FDR),]

```

